---
title: "class_5"
output: html_document
date: "2023-03-02"
---

```{r}
#load libraries
library(ISLR2)
library(leaps)
```

Let's continue using the 'Hitters' dataset: baseball player player performance stats, along with their salaries

```{r}
#omit NA
Hitters <- na.omit(Hitters)
```

# Subset selection

Let's perform best/forward/backward model selection
(These are methods that may improve our linear model fits, via feature selection)

```{r}

#perform best subset
#this suggests the variables in the best model (via smallest RSS) among all possible models with the *same number of predictors*
best_subset <- regsubsets(Salary ~ ., data=Hitters, nvmax=ncol(Hitters) -1) #we want all 19 vars considered

#look at summary
#note: asterisks indicate the variables included in the model at each possible model size (not significance!)
summary(best_subset)

#instead of 'best subset', let's perform forward stepwise selection
best_subset.fwd <- regsubsets(Salary ~ ., data=Hitters, nvmax=ncol(Hitters) -1, method='forward')

#let's perform backward stepwise selection
best_subset.bwd <- regsubsets(Salary ~ ., data=Hitters, nvmax=ncol(Hitters) -1, method='backward')

#when we compare the best/forward/backward approaches to model selection, we see that the variables included in each differ (this is ok/expected!)
```

We have not yet decided _which_ model, i.e., the number of predictors that is best (so far, we just know what variables/predictors may be included in models of various sizes, i.e., with a k value at each possible k size, from 1:19). We do this on the basis of estimating <u>test error.</u>

There are a number of competing ways to select the so-called best model. We discussed 4 "indirect" methods that adjust the test error, to avoid bias due to overfitting: `Cp`, `AIC`, `BIC` and `Adjusted $R^2$`.

```{r}

#let's save our summary, to more easily pull out values of interest
my_summary <- summary(best_subset)

#look at the adjusted R2 values -- one for each possible model size, i.e., each value of k
my_summary$adjr2

#we can make a nice plot (run this code all together)
plot(my_summary$adjr2, xlab='Number of Predictors', ylab='Adjusted R^2 value' ,type='l')
mypoint <- which.max(my_summary$adjr2) #highlight best value
points(mypoint, my_summary$adjr2[mypoint], col='red')

```
If we wanted, we can also select our ideal model via direct error estimation. This involves cross-validation. (We did not perform that in class, as similar to past and future examples).

# Shrinkage Methods

## Ridge regression
```{r}
#library
library(glmnet)

#the glmnet library requires our data in a special format
x <- model.matrix(Salary ~., Hitters)[, -1]
y <- Hitters$Salary

#split up data into training and testing (1/2 each)
set.seed(1)
train <- sample (1:nrow(x), nrow(x)/2)
test <- (-train)
y.test <- y[test]

#fit ridge regression model
ridge.mod <- glmnet(x[train, ], y[train], alpha = 0) #0=ridge

#cross validation -- helps us 'tune' lambda hyperparameter
#recall: we need to set lambda to help us achieve best fit
#note: the cv.glmnet function using nfolds=10
set.seed(1)
cv.ridge <- cv.glmnet(x[train, ], y[train], alpha = 0)

#we can make a nice plot, showing MSE, for each of the possible lambda values considered by cv.glmnet.
plot(cv.ridge)
#the top x axis values tell us number of non-0 predictors
#the vertical lines are the minimal lambda value that that largest within 1SE of the minimal value

#let's pull out our best minimal lambda 
(min1.lambda <- cv.ridge$lambda.min)

#note: there's a second estimate of the best lambda, as as largest lambda value within one SE of the minimal: helps avoid overfitted
(min2lambda <- cv.ridge$lambda.1se)

#now that we have the lambda with the smallest cross validation error, we can make our predictions in the test set
ridge.pred <- predict(ridge.mod, s=min1.lambda, newx = x[test, ])

#compute test MSE
(mse <- mean((ridge.pred - y.test)^2))

##################################################################
#NOTE
#In class, we noticed big discrepancy on the basis of seeds set
#To work out why this is, see separate `ridge_stability.Rmd'

#print out estimates of interest
min1.lambda
min2lambda
mse

#seed 123
#min1.lambda: 27.68026
#min2lambda: 1377.653
# mse: 145983.9

#seed 7
#min1.lambda: 29.5721
#min2lambda: 5413.877
#mse: 107880.1

#seed 1
#min1.lambda: 326.0828
#min2lambda: 6401.138
#mse: 139863.2

##################################################################

#after find ideal lambda, refit the ridge regression using the complete dataset
ridge <- glmnet(x, y, alpha = 0)

#pull out coefficient values (this is what we need to describe model)
predict(ridge, type = "coefficients", s = min1.lambda)

```

## Lasso

```{r}

#the process in R is virtually identical to that of ridge, with the exceptino that we now set alpha to 1
set.seed(123)
lasso.mod <- glmnet(x[train,], y[train], alpha=1)

#cross validation
cv.lasso <- cv.glmnet(x[train,], y[train], alpha=1)
plot(cv.lasso)

#find minimal lambda
lasso_min1.lambda <- cv.lasso$lambda.min

#make predictions, get MSE
lasso.pred <- predict(lasso.mod, s=lasso_min1.lambda, newx = x[test,])
mse <- mean((lasso.pred - y.test)^2)

#now, refit on full data, get coefficients
lasso <- glmnet(x, y, alpha=1)
predict(lasso, type = 'coefficients', s=lasso_min1.lambda)
```

